---
title: "Introduction to ggplot"
author: "Peter Freeman (2019 SLSW)"
date: "24 June 2019"
output: 
  slidy_presentation:
    font_adjustment: -1
---

Preliminaries
===

Let's read in the data frame that we use in the `dplyr` notes set:
```{r}
df = read.csv("http://www.stat.cmu.edu/~pfreeman/GalaxyMass.csv")
```
As a reminder, these data consist of 3456 rows and 10 columns with names
```{r}
names(df)
```

ggplot
===

`ggplot` (actually, and perhaps confusingly, `ggplot2`) is "a system for declaratively creating graphics, based on [The Grammar of Graphics](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448/ref=as_li_ss_tl?ie=UTF8&qid=1477928463&sr=8-1&keywords=the+grammar+of+graphics&linkCode=sl1&tag=ggplot2-20&linkId=f0130e557161b83fbe97ba0e9175c431). You provide the data [frame], tell `ggplot2` how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details."

Sounds good. Let's dive in:
```{r}
library(tidyverse)
```

ggplot: Basic Structure
===

A very basic call to `ggplot()` has the following structure:
```
ggplot(data=<data frame>,mapping=aes(x=<x axis variable>,...)) + geom_<plot type>(<arguments>)
```

ggplot: Bar Chart
===

How many galaxies are in each field?
```{r fig.height=6,fig.width=6,fig.align="center"}
ggplot(data=df,mapping=aes(x=field)) + geom_bar(color="red",fill="blue")
```
(A bar chart is appropriate when the x-axis variable is categorical and the y-axis variable is quantitative.) You can extract the same information in non-visual form using a call to `table()`:
```{r}
df %>% select(.,field) %>% table(.)
```

ggplot: Histogram
===

What is the distribution of galaxy sizes?
```{r fig.height=6,fig.width=6,fig.align="center"}
ggplot(data=df,mapping=aes(x=size)) + geom_histogram(color="blue",bins=25)
```
(A histogram is appropriate when the single variable in question is quantitative.)

ggplot: Boxplot
===

Boxplots are just a bit trickier. What is the distribution of galaxy concentrations?
```{r fig.height=6,fig.width=6,fig.align="center"}
ggplot(data=df,mapping=aes(x="",y=C)) + geom_boxplot(color="blue",fill="orange")
```
Note the empty string for `x`.
(A boxplot is also for visualizing a quantitative variable.)

ggplot: Side-By-Side Boxplots
===

We can group the data given the values of a categorical variable to create side-by-side boxplots:
```{r fig.height=6,fig.width=6,fig.align="center"}
ggplot(data=df,mapping=aes(x=field,y=C)) + geom_boxplot(color="blue",fill="orange") 
```

ggplot: Faceting
===

Faceting is the act of making multiple plots at once that appear side-by-side as "facets". Faceting is something you might want to do when, e.g., you have a factor variable. Here, we show histograms of the concentration variable `C` broken up by galaxy field.
```{r fig.height=6,fig.width=6,fig.align="center"}
ggplot(data=df,mapping=aes(x=C)) + geom_histogram(color="blue",fill="yellow",bins=25) +
facet_wrap(~field,scales='free_x')
```

ggplot: Scatter Plot
===

Is there any visually apparent association between the mass of a galaxy and the concentration of its light?
```{r fig.height=6,fig.width=6,fig.align="center"}
ggplot(data=df,mapping=aes(x=mass,y=C)) + geom_point(color="red",size=0.5)
```

Covariance and Correlation
===

Covariance is a measure of the *linear* dependence between two variables. (To be "uncorrelated" is not the same as to be "independent"...the latter means there is no dependence, linear or otherwise, between two variables.) Correlation is a "normalized" form of covariance, that ranges from -1 (one variable linearly decreases absolutely in value while the other increases in value) through 0 (no linear dependence) to 1 (one variable linear increases absolutely while the other increases).

A good package for visualizing correlation is `corrplot`:
```{r}
if ( require(corrplot) == FALSE ) {
  install.packages("corrplot",repos="https://cloud.r-project.org")
  library(corrplot)
}
```

corrplot: Correlation Plot
===

Here we generate the correlation plot for `M20`, `A`, and `z.mode`:
```{r fig.height=6,fig.width=6,fig.align="center"}
corrplot(cor(df[,c("M20","A","z.mode")]),method="ellipse")
```

GGally: Pairs Plot
===

We step outside canonical `ggplot` plotting for now to bring you a pairs plot from the `GGally` package:
```{r fig.height=6,fig.width=6,fig.align="center"}
if ( require(GGally) == FALSE ) {
  install.packages("GGally",repos="https://cloud.r-project.org")
  library(GGally)
}
df.new = df %>% filter(.,(field=="COSMOS"|field=="EGS")) %>% select(M20,A,z.mode)
ggpairs(df.new,progress=FALSE,lower=list(combo=wrap("facethist", binwidth=0.8)))
```

